\protect Connecting every bit of knowledge: \\ The structure of Wikipedia's first link network

### 2016-03-09 writeup/wikipedia-network.tex
chris suggestions incorporated

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/articles_ndegree.png}
  \caption{
    \textbf{Highest Ranking Articles by in-degree.}
    We rank each article by the number of direct first links to the article (in-degree). The highest-ranking articles tend to represent geographical and biological abstractions. A full online appendix of the results and data is available \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.}
  \label{fig:indegree list}
\end{figure}

@@

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/ndegree_loglog.png}
  \caption{
    \textbf{FLN Degree Distribution.}
We fit the in-degree distribution with a linear $\log_{10}-\log_{10}$ model. 
The result is a 
an excellent fit with a Pearson's Correlation Coefficient of $-0.98$, yielding a 
power law rank exponent $\alpha = -0.788$ and a size exponent $\gamma = -0.266$. 
}
  \label{fig:degree distribution}
\end{figure}

@@

Of the 11 million articles, 5.5 million had an invalid link or linked back to the same article, yielding a path length of zero. 
This roughly corresponds to the official number of articles on Wikipedia: 
$~4.7$ million as of November 2014---approximately half of the 11 million 
articles in the XML dump are redirects or disambiguations, not full articles.
The most common path length is 29, with an interquartile range of 4---path length
is typically between 26 and 29 articles.
As a distribution, more than $75\%$ of articles have a path length below 
$50$ first links 
while a few temporally organized paths exceed 50 links 
(see Fig.~\ref{fig:Path Length Distribution}).

@@

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/traversals_per_article.png} 
  \caption{
    \textbf{Distribution of Traversal Visits.}
We fit the distribution of traversal visits against article rank with a log-log (base 10) linear regression model log-log model and uncover two regimes: 
with a cutoff rank of $10^{5}$.
The top regime ($\log_{10}$(rank) $< 5$) has a Pearson's Correlation 
Coefficient of $-0.99$ 
with a corresponding power law rank exponent 
$\alpha = -1.23$ and a size exponent $\gamma = 0.187$.
The horizontal flattening around the highest
ranking articles is a result of the cyclic structure (see discussion on cycles).
The bottom regime has a Pearson's Correlation Coefficient of $-0.93$
with a corresponding power law rank exponent 
$\alpha = -0.579$ and a size exponent $\gamma = -0.727$.}
  \label{fig:Distribution of Visits}

In Fig.~\ref{fig:Distribution of Visits}, we show traversal visits against rank, yielding
an excellent fit with regimes corresponding to power law rank exponents of 
$\alpha = -1.23 (\gamma = 0.187$) and $\alpha = -0.579 (\gamma = -0.727)$.
On the aggregate, the distribution suggests
a handful of the highest ranking articles contain a disproportionate number of traversal visits, while most have none. The skew in the distribution is not terribly surprising when considering the heuristic of how the links flow: from specific to general.

@@

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{graphics/2_cycles.pdf}
  \caption{
    \textbf{Highest ranking 2-cycles.}
We identify pairs of articles whose first links point to one another, forming
a 2-cycle. We then rank each pair of articles by the total number of 
traversal visits to gauge the most referenced groups of two articles linked
to each other. We find 2-cycles often capture synonyms or articles representing nearly the 
same concepts as opposed to distinct concepts.}
  \label{fig:2-cycles}
\end{figure*}
\subsection{Network Cycles}
\label{cycles}

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{graphics/3_cycles.pdf}
  \caption{
\textbf{Highest ranking 3-cycles.}
We identify groups of three articles whose links form a 3-cycle with the 
last article in the group linking back to the first article.
We rank each 3-cycle by the total number of traversal visits to gauge
the most referenced group of three linked articles. 
The area of the circles in the figure indicates the number of traversal visits.}
  \label{fig:3-cycles}
\end{figure*}
We identify 2-cycles, meaning a pair of articles with first link pointing to one another.
Of the 11 million articles, roughly 84,000 are memebers of 2-cycles. 
The highest ranking 2-cycles by traversal visits tend to be synonyms (or nearly so) rather than distinct, yet connected topics:
``Health Care'' and ``Medical Case Management,'' ``Broadcasting House,'' and ``BBC,'' ``Secondary Education'' and ``Secondary School'' 
(see Fig.~\ref{fig:2-cycles}).

@@

@@

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/top_funnels.png}
  \caption{
    \textbf{Funnels.}
We represent the highest-ranking articles by the number of traversal 
funnels to gauge the influence each article exerts in shaping the 
structure of the FLN. 
We find ``Philosophy'' exerts an overwhelming proportion
of the influence, with other abstract notions and topical concepts ranking
next.}
  \label{fig:Funnels}
\end{figure}
To analyze the influence an article exerts in shaping the 
structure of the FLN, we compute the number of traversal funnels for each.
Articles directing more paths exert a greater influence over the structure
of the FLN by increasing the accumulation of first links
on a particular path. By measuring traversal funnels, we distinguish between an article that simply happened to fall within a cycle from an article funneling 
many first links.

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/funnels_distribution.png}
  \caption{
    \textbf{Distribution of Traversal Funnels.}
We fit the distribution with a linear log-log model by considering the log (base 10) transformed rank of each article against log (base 10) transformed number of traversal funnels. 
We find two regimes with the top regime (log(rank) < 4) 
well-explained by a linear fit, yielding Pearson's 
Correlation Coefficient of $-0.99$ and a 
corresponding power law rank exponent $\alpha = -1.08$
and a size exponent $\gamma = 0.074$. 
The bottom horizontal regime corresponds to the more than $99\%$ of articles
which hold zero traversal funnels.}
  \label{fig:Funnels Distribution}
\end{figure}
Ranking articles by the number of traversal funnels we find 
``Philosophy'' to be by far the highest-ranking article with 
$7.37$ million paths
(see Fig.~\ref{fig:Funnels}).
Of any article, the number of traversal funnels Philosophy holds exceeds 
all others by more than two orders of magnitude.
The ``Philosophy'' cycle which contains ``Existence,'' ``Awareness,'' ``Reality,'' 
and similar articles accumulates the overwhelming proportion of its 
references through ``Philosophy'': $7.37$ million of the $7.4$ million references
are funneled through ``Philosophy''.
Second on the list of highest-ranking articles by traversal funnels is 
``Presentation'' with only $30$ thousand paths. Similarly abstract 
concepts also rank highly such as ``Tree of life'' (30 thousand), 
``Reality'' (13 thousand), and ``Jurisdiction'' (3 thousand).

@@

As a distribution, we find few articles influence the structure of the 
FLN. Only $17, 821$ articles have one or more traversal funnels, leaving
more than $99\%$ with none---most articles are recipients of 
the references flowing through the articles with at least one traversal funnel.
When fit to a log-log linear model we find the $99\%$ of articles with zero
traversal funnels form one regime (with log(rank) less than 4).
The top regime, corresponding to the $17, 821$ articles with at least one 
traversal funnel strongly fits a linear model with
a Pearson's Correlation Coefficient of $-0.99$ and a 
corresponding power law rank exponent $\alpha = -1.08$
and a size exponent $\gamma = 0.074$ 
(see Fig.~\ref{fig:Funnels Distribution}). Even within the few articles
influencing the structure of the FLN, only a handful of these exert most of the 
control.

\subsection{Article Popularity}

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/views_visits.png}
  \caption{
\textbf{Article Popularity (by page views) compared to Traversal Visits 
for Highest ranking 1000 articles.}
We use the total number of page views provided by Wikipedia for the month
of October 2015 to compare each article's popularity to traversal visits.
While the most popular articles do not necessarily correspond to the articles
with the highest number of traversal visits, the variation in popularity appears to decrease as the number of traversal visits increases. The greatest number of articles fall within roughly 1 million traversal visits and page views.}
  \label{fig:Views and Visits}

\end{figure}
Wikipedia released an API to measure article popularity by page views
starting November 2015. 
Page views add another dimension to our
findings by constrasting
the number of users who access a particular article against
the number of traversal visits and funnels.
We measure popularity as the total number 
of page views in the English edition of Wikipedia in the month of 
October 2015---the earliest full month for which the data is available. 
We find the highest ranking 1000 articles (by traversal visits) have an average of
$70$ thousand page views in October with high variation: the standard deviation 
is $1.1$ million page views. 
The number of page views has a skewed distribution with $75\%$ of articles
reaching fewer than $73$ thousand page views in October.
The article for ``United States'' has the most page 
views of the highest ranking $1000$ articles by traversal visits with 
$1.4$ million views in October. The next most popular articles are 
``India,'' ``World War II,'' and the ``United Kingdom'' each with roughly 1 million page views. 
The ``Philosophy'' article, despite outranking every article by traversal visits,
has only $240$ thousand page views in October.

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/funnels_visits.png}
  \caption{
    \textbf{Article Popularity (by page views) compared to Traversal Funnels
    for Highest ranking 1000 articles.}
We use the total number of page views provided by Wikipedia for the month
of October 2015 to compare each article's popularity to traversal funnels.
In the top left is ``Halloween'' with a significantly greater 
number of page views in the month of October. 
The greatest number of articles falls within $10^3-10^5$ page views, 
but only $10^{1.2}-10^2$ traversal funnels.}
  \label{fig:Views and Funnels}
\end{figure}

@@

To process the entirety of Wikipedia, we distributed the parsing and processing of the XML dump
across 112 cores of the UVM supercomputer cluster
\cite{vacc}.
We then joined the results to form a hash table containing every Wikipedia article and its corresponding
first link. The resulting network map is the basis of our analysis.

An online appendix containing all of the code and data used can be accessed \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.

>>>>

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/articles_ndegree.png}
  \caption{
    \textbf{Highest Ranking Articles by in-degree.}
    We rank each article by the number of direct first links to the article (in-degree). The highest-ranking articles tend to represent geographical and biological abstractions. A full online appendix of the results and data is available \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.}
  \label{fig:indegree list}
\end{figure}
\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/ndegree_loglog.png}
  \caption{
    \textbf{FLN Degree Distribution.}
We fit the in-degree distribution with a linear $\log_{10}-\log_{10}$ model. 
The result is a 
an excellent fit with a Pearson's Correlation Coefficient of $-0.98$, yielding a 
power law rank exponent $\alpha = -0.788$ and a size exponent $\gamma = -0.266$. 
}
  \label{fig:degree distribution}
\end{figure}

@@

@@

Of the 11 million articles, 5.5 million had an invalid link or linked back to the same article, yielding a path length of zero. 
\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/traversals_per_article.png} 
  \caption{
    \textbf{Distribution of Traversal Visits.}
We fit the distribution of traversal visits against article rank with a log-log (base 10) linear regression model log-log model and uncover two regimes: 
with a cutoff rank of $10^{5}$.
The top regime ($\log_{10}$(rank) $< 5$) has a Pearson's Correlation 
Coefficient of $-0.99$ 
with a corresponding power law rank exponent 
$\alpha = -1.23$ and a size exponent $\gamma = 0.187$.
The horizontal flattening around the highest
ranking articles is a result of the cyclic structure (see discussion on cycles).
The bottom regime has a Pearson's Correlation Coefficient of $-0.93$
with a corresponding power law rank exponent 
$\alpha = -0.579$ and a size exponent $\gamma = -0.727$.}
  \label{fig:Distribution of Visits}

This roughly corresponds to the official number of articles on Wikipedia: 
$~4.7$ million as of November 2014---approximately half of the 11 million 
articles in the XML dump are redirects or disambiguations, not full articles.
The most common path length is 29, with an interquartile range of 4---path length
is typically between 26 and 29 articles.
As a distribution, more than $75\%$ of articles have a path length below 
$50$ first links 
while a few temporally organized paths exceed 50 links 
(see Fig.~\ref{fig:Path Length Distribution}).

@@

\begin{figure*}[tph!]
  \includegraphics[width=\textwidth]{graphics/2_and_3_cycles.pdf}
  \caption{
    \textbf{Highest ranking 2-cycles and 3-cycles.}
We identify pairs of articles whose first links point to one another, forming
a 2-cycle. We then rank each pair of articles by the total number of 
traversal visits to gauge the most referenced groups of two articles linked
to each other. We find 2-cycles often capture synonyms or articles representing nearly the 
same concepts as opposed to distinct concepts. Similarly, we identify and 
rank 3-cycles to find they appear to capture three closely related ideas or synonyms.}
  \label{fig:2-cycles}
\end{figure*}
In Fig.~\ref{fig:Distribution of Visits}, we fit traversal visits against rank
to two regimes corresponding to power law rank exponents of 
$\alpha = -1.23 (\gamma = 0.187$) and $\alpha = -0.579 (\gamma = -0.727)$.
On the aggregate, the distribution suggests
a handful of the highest ranking articles contain a disproportionate number of traversal visits, while most have none. The skew in the distribution is not terribly surprising when considering the heuristic of how the links flow: from specific to general.

@@

\subsection{Network Cycles}
\label{cycles}
\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/top_funnels.png}
  \caption{
    \textbf{Funnels.}
We represent the highest-ranking articles by the number of traversal 
funnels to gauge the influence each article exerts in shaping the 
structure of the FLN. 
We find ``Philosophy'' exerts an overwhelming proportion
of the influence, with other abstract notions and topical concepts ranking
next.}
  \label{fig:Funnels}
\end{figure}

We identify 2-cycles, meaning a pair of articles with first link pointing to one another.
Of the 11 million articles, roughly 84,000 are memebers of 2-cycles. 
The highest ranking 2-cycles by traversal visits tend to be synonyms (or nearly so) rather than distinct, yet connected topics:
``Health Care'' and ``Medical Case Management,'' ``Broadcasting House,'' and ``BBC,'' ``Secondary Education'' and ``Secondary School'' 
(see Fig.~\ref{fig:2-cycles}).

@@

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/funnels_distribution.png}
  \caption{
    \textbf{Distribution of Traversal Funnels.}
We fit the distribution with a linear log-log model by considering the log (base 10) transformed rank of each article against log (base 10) transformed number of traversal funnels. 
We find two regimes with the top regime (log(rank) < 4) 
well-explained by a linear fit, yielding Pearson's 
Correlation Coefficient of $-0.99$ and a 
corresponding power law rank exponent $\alpha = -1.08$
and a size exponent $\gamma = 0.074$. 
The bottom horizontal regime corresponds to the more than $99\%$ of articles
which hold zero traversal funnels.}
  \label{fig:Funnels Distribution}
\end{figure}

@@

To analyze the influence an article exerts in shaping the 
structure of the FLN, we compute the number of traversal funnels for each.
Articles directing more paths exert a greater influence over the structure
of the FLN by increasing the accumulation of first links
on a particular path. By measuring traversal funnels, we distinguish between an article that simply happened to fall within a cycle from an article funneling 
many first links.

Ranking articles by the number of traversal funnels we find 
``Philosophy'' to be by far the highest-ranking article with 
$7.37$ million paths
(see Fig.~\ref{fig:Funnels}).
Of any article, the number of traversal funnels Philosophy holds exceeds 
all others by more than two orders of magnitude.
The ``Philosophy'' cycle which contains ``Existence,'' ``Awareness,'' ``Reality,'' 
and similar articles accumulates the overwhelming proportion of its 
references through ``Philosophy'': $7.37$ million of the $7.4$ million references
are funneled through ``Philosophy''.
Second on the list of highest-ranking articles by traversal funnels is 
``Presentation'' with only $30$ thousand paths. Similarly abstract 
concepts also rank highly such as ``Tree of life'' (30 thousand), 
``Reality'' (13 thousand), and ``Jurisdiction'' (3 thousand).

@@

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/popularity_visits.pdf}
  \caption{
\textbf{Article Popularity (by page views) for the highest ranking 1000 articles by traversal visits.}
We use the total number of page views provided by Wikipedia for the month
of October 2015 to compare each article's popularity to traversal visits.
While the most popular articles do not necessarily correspond to the articles
with the highest number of traversal visits, the variation in popularity appears to decrease as the number of traversal visits increases. The greatest number of articles fall within roughly 1 million traversal visits and page views. Overall, the page views for the top 1000 articles articles by traversal visits appear to be log-normally distributed. 
}
  \label{fig:Views and Visits}

\end{figure}
As a distribution, we find few articles influence the structure of the 
FLN. Only $17, 821$ articles have one or more traversal funnels, leaving
more than $99\%$ with none---most articles are recipients of 
the references flowing through the articles with at least one traversal funnel.
When fit to a log-log linear model we find the $99\%$ of articles with zero
traversal funnels form one regime (with log(rank) less than 4).
The top regime, corresponding to the $17, 821$ articles with at least one 
traversal funnel strongly fits a linear model with
a Pearson's Correlation Coefficient of $-0.99$ and a 
corresponding power law rank exponent $\alpha = -1.08$
and a size exponent $\gamma = 0.074$ 
(see Fig.~\ref{fig:Funnels Distribution}). Even within the few articles
influencing the structure of the FLN, only a handful of these exert most of the 
control.

\begin{figure}[tp!]
  \includegraphics[width=\columnwidth]{graphics/popularity_funnels.pdf}
  \caption{
\textbf{Article Popularity (by page views) for the highest ranking 1000 articles by traversal funnels.}
We use the total number of page views provided by Wikipedia for the month
of October 2015 to compare each article's popularity to traversal funnels.
}
  \label{fig:Views and Funnels}
\end{figure}
\subsection{Article Popularity}

Wikipedia released an API to measure article popularity by page views
starting November 2015. 
Page views add another dimension to our
findings by constrasting
the number of users who access a particular article against
the number of traversal visits and funnels.
We measure popularity as the total number 
of page views in the English edition of Wikipedia in the month of 
October 2015---the earliest full month for which the data is available. 
We find the highest ranking 1000 articles (by traversal visits) have an average of
$70$ thousand page views in October with high variation: the standard deviation 
is $1.1$ million page views. 
The number of page views has a skewed distribution with $75\%$ of articles
reaching fewer than $73$ thousand page views in October.
The article for ``United States'' has the most page 
views of the highest ranking $1000$ articles by traversal visits with 
$1.4$ million views in October. The next most popular articles are 
``India,'' ``World War II,'' and the ``United Kingdom'' each with roughly 1 million page views. 
The ``Philosophy'' article, despite outranking every article by traversal visits,
has only $240$ thousand page views in October.

@@

To process the entirety of Wikipedia, we distributed the parsing and processing of the XML dump
across 112 cores of the UVM supercomputer cluster
\cite{vacc}.
We then joined the results to form a hash table containing every Wikipedia article and its corresponding
first link. The resulting network map is the basis of our analysis.
An online appendix containing all of the code and data used can be accessed \href{http://compstorylab.org/share/papers/ibrahim2016a/index.html}{here}.
